Developing an AI-Assisted Grading System Using Large Language Models (GPT-4)

### 2024-11-18 src/paper.tex
Dr. A's final edits

% Include necessary packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[sc,osf]{mathpazo}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{color}
\definecolor{dark-red}{rgb}{0.6,0,0}
\definecolor{dark-green}{rgb}{0,0.6,0}
\definecolor{dark-blue}{rgb}{0,0,0.6}

@@

% Title and author information
\title{Developing an AI-Assisted Grading System Using Large Language Models (GPT-4)}
\author{Andrei Modiga}
\adviser{Professor Anderson}
\adviserAbstract{Scot Anderson, Ph.D.}
\major{Computer Science}
\degreemonth{August}
\degreeyear{2024}

@@

The primary goal of this project is to build and implement this AI-assisted grading system, including the ability to recognize handwritten text, and integrate it into an educational platform like OICLearning.com. By focusing on building a practical solution, the project aims to enhance the grading process, making it quicker and more consistent, thereby allowing teachers to focus on more critical aspects of instruction while ensuring that students receive fair and accurate feedback.

@@

\begin{itemize}
    \item \textbf{Backend API Development:} Implement a backend API in Python that can receive images and text submissions from students and send them to GPT-4 and GPT-4 Vision for processing.
    \item \textbf{Integration with GPT-4 Vision:} Utilize GPT-4 Vision's capabilities to perform OCR and NLP on images of student submissions.
    \item \textbf{Grading Algorithm:} Develop algorithms to compare the processed student responses against expected answers and assign appropriate grades.
    \item \textbf{Frontend Interface:} Create a user-friendly frontend interface that allows instructors to view, sort, and manage graded assignments.
    \item \textbf{Data Security and Privacy:} Ensure the system complies with data protection regulations, implementing necessary security measures.
    \item \textbf{Scalability:} Design the system to handle a large number of student submissions and various types of assignments.
    \item \textbf{User Authentication:} Implement secure login and access control mechanisms for instructors.
\end{itemize}

@@

\begin{itemize}
    \item Receive images and text submissions from the frontend.
    \item Send images to GPT-4 Vision via API calls for OCR and NLP processing.
    \item Parse and handle responses from GPT-4.
    \item Compare the interpreted responses to predefined expected answers.
    \item Calculate grades based on a grading rubric or algorithm.
    \item Store the results in a database.
\end{itemize}

@@

The frontend will be a web-based application that allows instructors to:

\begin{itemize}
    \item Securely log in to the system.
    \item View graded assignments and student responses.
    \item Sort and filter submissions based on various criteria (e.g., grade, date submitted).
    \item Override grades or make manual adjustments.
\end{itemize}

@@

@@

\begin{itemize}
    \item Fully functional AI-assisted grading system integrated with OICLearning.com.
    \item Source code for both backend and frontend components.
    \item Technical documentation detailing system architecture, setup, and usage.
    \item User guide for instructors on how to use the system.
    \item Test cases and testing reports.
    \item Deployment scripts and configuration files.
\end{itemize}

\section{Development Prerequisites}

The development of this AI-assisted grading system requires the following:

\begin{itemize}
    \item Access to OpenAI's GPT-4 and GPT-4 Vision APIs.
    \item Development environment set up with Python 3.x.
    \item Web server and hosting for the frontend application.
    \item Secure database server for storing data.
    \item Knowledge of data protection regulations applicable to student data.
\end{itemize}

\appendix

\chapter{Requirements Specification}

@@

\backmatter

>>>>

% Include necessary packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[sc,osf]{mathpazo}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{color}
\definecolor{dark-red}{rgb}{0.6,0,0}
\definecolor{dark-green}{rgb}{0,0.6,0}
\definecolor{dark-blue}{rgb}{0,0,0.6}
\interfootnotelinepenalty=10000 % This helps to prevent breaking footnotes across pages

@@

% Title and author information
\title{Developing an AI-Assisted Grading System Using Large Language Models (GPT-4)}
\author{Andrei Modiga}
\adviser{Scot Anderson, Ph.D.}
\adviserAbstract{Scot Anderson, Ph.D.}
\major{Computer Science}
\degreemonth{August}
\degreeyear{2024}

@@

The primary goal of this project is to build and implement this AI-assisted grading system, including the ability to recognize handwritten text, and integrate it into the educational platform, OICLearning.com. By focusing on building a practical solution, the project aims to enhance the grading process, making it quicker and more consistent, thereby allowing teachers to focus on more critical aspects of instruction while ensuring that students receive fair and accurate feedback.

@@

\begin{itemize}
    \item \textbf{Backend API Development:} Implement a backend API in Python that can receive images and text submissions from students and send them to GPT-4 and GPT-4 Vision for 
    processing.
    %% Summary - this is covered in the previous point. 
    %\item \textbf{Integration with GPT-4 Vision:} Utilize GPT-4 Vision's capabilities to perform OCR and NLP on images of student submissions.
    % \item \textbf{Grading Algorithm:} Develop algorithms to compare the processed student responses against expected answers and assign appropriate grades.
    \item \textbf{Frontend Interface:} Create a user-friendly frontend interface that allows instructors to view, sort, and manage the process of grading assignments.
    \item \textbf{Data Security and Privacy:} Ensure the system complies with data protection regulations, implementing necessary security measures.
    \item \textbf{Scalability:} Design the system to handle a large number of student submissions and various types of assignments.
    \item \textbf{User Authentication:} Integrate with the secure login and access control mechanisms avaialable in the OICLearning.com project.
\end{itemize}

@@

\begin{itemize}
    \item Receive images and text submissions from the frontend.
    \item Send images to GPT-4 Vision via API calls for OCR and NLP processing.
    \item Parse and handle responses from GPT-4.
    \item Compare the interpreted responses to predefined expected answers. %Dr. A says, OK this may be needed to do accurate grouping.
    \item Store the results in a database.
\end{itemize}

@@

The frontend will be a web-based application that allows users to:

\begin{itemize}
    \item Securely log in to the system.
    \item Upload and submit assignments. %Added by Dr. A
    \item View graded assignments and student responses.
    \item Sort and filter submissions based on various criteria (e.g., grade, date submitted).
    \item Calculate grades based on a grading rubric or algorithm.
    \item Override grades or make manual adjustments.
\end{itemize}

@@

\section{Detailed Requirements and User Stories}

Four user stories describe the processes that drive the requirements. These stories include: the teacher creating an assignment, teachers or students uploading submissions, the AI assisted grading of assignments and finally the teacher transfering the grades to an Learning Management system (LMS). These stories and their associated requirements describe a series of webpages shown in Figure~\ref{fig:page-flow}  The following sections detail each of these stories.

\begin{figure}[h]
    \centering
    \includegraphics[width=.95\textwidth]{images/page-flow.pdf}
    \caption{Page flow diagram showing the sequence of pages in the system.}
    \label{fig:page-flow}
\end{figure}

\subsection{A teacher creates an assignment}

When a teacher creates an assignment, they choose from different kinds of AI assisted types. These types include:

\begin{itemize}
    \item Filled-Form that students write on and are later scanned in by the teacher. E.g. a written test. 
    \item Free-Form typed or written assignments where the location of answers may vary. E.g. an assignment that is given to the students in a word document and the length of their answer may vary, and hence their location may vary.
    \item Online Filled-form assignments. E.g. a quiz that is taken online where answers that contains digital responses from students. This is not within the scope of this project.
\end{itemize}

For Filled-Form type assignments, the teacher needs to prompt the system by uploading a blank of the form. Figure~\ref{fig:filled-form} shows an example of what this page might look like. The teacher first identifies the location of the student identity (name and/or ID number). In the Figure, this is shown by the tan box added by the teacher around the area where a student is instructed to place their name. To add a question, the teacher clicks on the ``New Question'' link at the bottom right. The system adds a new question to the list and an associated box to go with it. Then the teach must: name the question, give the points , locate and size the associated box (shown in green) of the answer area over the blank spot they expect students to write in, and give the expected answer (not shown). Figure~\ref{fig:filled-form} shows the name area and a single question where the teacher can accomplish all these things. Here when a teacher clicks on a question in the list, they can edit the box size/location along with the name or points. The plus sign next to the red X, allows the teacher to add a sub-question. Sub-questions work the same way as questions except they appear as a sublist under the question.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{images/filled-form-creation.png}
    \caption{The filled-form assignment creation page.}
    \label{fig:filled-form}
\end{figure}

This process is called editing the outline and includes everything the system needs to define the assignment. The teacher can then save the outline and the system will be ready to accept student submissions. The teacher can create rubrics at this point or, more likely, defer this step grading. If the teacher decides to create the rubrics now, they progress to the webpage called ``Create Rubrics'' (see Figure~\ref{fig:page-flow}). Otherwise, they can add rubrics during grading.

A rubric defines a value subtracted from a score when applied, and a description of the reason for the deduction. The teacher can add up to 36 rubrics and they shall be numbered 0-9, and a-z. Teachers assign one or more rubrics to an answer (or group of answers when grading groups) during grading.

In the case of Free-formed typed or written assignments, the teacher gives the students questions, and the students turn in their answers as a PDF. Students may or may not include the questions, so the answer's location will vary from student to student. Here the teacher does not need to identify the location of an answer, but they do need to provide question identifiers and points for students to use when turning in the document. Figure~\ref{fig:free-form} shows an example of the teacher creating a Free-Form assignment. Notice that there is no need to draw a rectangle around the answer area and except for that task the teacher's work is identical to the filled-form assignment above.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{images/free-form-creation.png}
    \caption{The free-form assignment creation page.}
    \label{fig:free-form}
\end{figure}

\subsection{Preparing Submissions to be graded}

The two tasks at this stage are uploading submissions and assigning names to them. This is done in two steps. First, the assignments are uploaded by the teacher or by the individual student. Second, the teacher approves/assigns names to the assignments. This takes two screeens Manage Scans and Manage Submissions, or one screen when uploaded by the student (refer to the pages list in Figure~\ref{fig:page-flow}).

\textit{Manage scans}: After a teacher has logged in and gone to a filled-in assignment, they upload all the assignments as a single PDF. The system already knows how many pages are in each assignment because it identified that number in the creation process above. The system separates the PDF into individual documents and attempts to assign the names of students based on the identity. This is accomplished on a screen called Manage. However, this page does not show the individual assignments or the names it has assigned. It only shows the number of assignments it assigned~\footnote{\scriptsize{In other grading systems, PDFs are converted to images. Although we would like to deal with PDFs natively, this is an acceptable solution}}.

\text{Manage Submissions}: In this page, a teacher can view a portion of each submission that contains the identity of the student. The system knows the students in the class and populates drop-down selection boxes for the names of students that it could not identify. The teacher can then select the correct student name from the drop-down. Of course, this is only necessary for Filled-form assignments uploaded by the teacher. Free-form assignments that students upload are automatically assigned to the student that uploaded them. Consequently, Free-form assignments will not have this page in the list of pages teachers see to manage the assignment.

\subsection{Grading Submissions}

The process starts with a list of questions for an assignment that needs grading. Each question is a link to a page where grouping can occur. The AI assists in this process by default, but the teacher can override the AI's grouping, or stop the process and manually group answers.

In the AI assisted process, the screen shows the list of groups on the right with the main screen showing the student answers as images in a grid. Figure~\ref{fig:answer-grouping} shows an example with eight answers. The first three in the left column have been selected by clicking on the images, and at this point would form a new group if the user clicks on the ``+Create a Group'' button. The AI should group the results first, but may leave some in an ungrouped ``group.'' That is, single submissions that don't fit into a group need not be put in a separate group, but should be kept in an ``ungrouped'' group. \textit{All} groups should be listed at the right. When clicking on the group, the system should de-select any images currently selected and then show the newly selected group's images in the main page. When images are selected and the ``Create a Group'' button is clicked, the system should create the group and move the images from the current group to the new group, then display the new group and it's images.

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.95\textwidth]{images/grouping-answers.png}
    \caption{Grouping answers.}
    \label{fig:answer-grouping}
\end{figure}

Selecting a group shows the group's Information. This information includes the images in the group (in the main screen), the name of the group, the number of images in the group, and a description of the group. All of these are assigned by the AI and the teacher may edit these items as appropriate.

Before moving to the next step it is important that the teacher review all the groups and that they are satisfied that the groups are correct. If the teacher decides that the grouping is incorrect, they must come back to this page to regroup them.

Once the system has groups created, the teach can proceed to grade the answers. The teacher does this by clicking on the Grade Answers button at the lower right in Figure~\ref{fig:answer-grouping}. Groups are graded in order that they appear on the screen with the sole exception that no matter where the ungrouped group is, it is always graded last.

When grading a group, the teacher is shown the group name, the number of images in the group, a sample answer(s) from the group, and the description of the group. The teacher can then assign rubrics that subtract points and provide feedback to students. The description of the group is part of the feedback that the student will see. Figure~\ref{fig:grading-answers} shows and example of what the grading page might look like. Several important requirements come into play for ease of use, specifically key presses and mouse movements have the following meaning:

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.95\textwidth]{images/grading-answer.png}
    \caption{The grading page.}\label{fig:grading-answers}
\end{figure}

\begin{itemize}
    \item Pressing 0--9 or some key combination with a--z toggles the rubric as assigned or not-assigned to the group. 
    \item Left arrow takes the teacher to the previous group or individual submission.
    \item Right arrow takes the teacher to the next group or individual submission.
    \item Scrolling up or down over the rubrics scrolls through the available rubrics.
    \item Scrolling on the image zooms in and out.
    \item Clicking and dragging the image moves it around in the viewing area. 
\end{itemize}

Suppose a teacher decides to change the value of a rubric. They can simple edit the rubric while grading and all grades where this rubric applies are updated with the new value.

When the teacher completes all of the groups and each individual assignment in the ``ungroupped'' group, they can move on to the next question in the assignment. When they are finished with all the questions, they can publish the results so that students can see them. This ``Review Grades'' page allows the teacher to see a list of all the grades has buttons to do the following actions:

\begin{itemize}
    \item Publish or Unpublish grades (Students see grades or not)
    \item Download Grades (export grades to an Excel or CSV file)
    \item Export submissions (to a zip file). Submissions are in PDF format and include the following:
    \begin{itemize}
        \item Name and Identifying information of the student along with the assignment score.
        \item A list of question names with the points assigned and the rubrics applied to that question. 
        \item After the grade information, The students original work is shown.
    \end{itemize}
    \item Export Evaluations (to a zip file of CSV files see Figure~\ref{fig:evaluation-export} for an example of a single question evaluation). 
\end{itemize}

\begin{figure}[htb]
    \centering
    \includegraphics[width=\textwidth]{images/evaluation-exports.png}
    \caption{Comma separated values (CSV) file example.}\label{fig:evaluation-export}
\end{figure}

This naturally brings us to the last step, transferring grades to an LMS.

\subsection{Transferring Grades to an LMS}

For the time being, this means downloading an Excel or CSV file that includes the student identifying information and the grade (e.g. 11/15) and a percentage (e.g. 73) without the percent sign. The teacher can then upload this file to the LMS.

@@

\begin{itemize}
    \item Fully functional AI-assisted grading system integrated with OICLearning project.
    \item Source code for both backend and frontend components.
    \item Technical documentation detailing system architecture, setup, and usage.
    \item User guide for instructors on how to use the system.
    \item Test cases and testing reports.
    \item Deployment scripts and configuration files.
\end{itemize}

% \section{Development Prerequisites}

% The development of this AI-assisted grading system requires the following:

% \begin{itemize}
%     \item Access to OpenAI's GPT-4 and GPT-4 Vision APIs.
%     \item Development environment set up with Python 3.x.
%     \item Web server and hosting for the frontend application.
%     \item Secure database server for storing data.
%     \item Knowledge of data protection regulations applicable to student data.
% \end{itemize}

%\appendix

%\chapter{Requirements Specification}

@@

%\backmatter
